\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-28 tex/paper.tex
save

In this note, we develop a likelihood-free inference algorithm for the point
estimation from observed data of the parameters of a forward non-differentiable
generative model. We propose to adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

Evolution strategies~\citep{2011arXiv1106.4487W} and variational optimization~\cite{2012arXiv1212.4507S} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
\end{equation}
where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$ and .
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

@@

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. On the other hand, for small values of $\gamma$,
proposal distributions with larger entropy are not penalized, which may result
in learning a smeared variation of the original simulator.
\glnote{Add note on minimum entropy, otherwise it degenerates back to a non-differentiable function.}

@@

For continuous data, we first consider inference for
a univariate Gaussian with unknown mean and
unknown variance. The observed data is sampled from a univariate Gaussian with
mean $\mu = 1$ and variance $\sigma^2 = 0.5^2$. Algorithm~\ref{alg:avo} is run
for 300 epochs with mini-batches of size $M=64$ and the following configuration.
For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
first and second moment vectors reset at each epoch in order to avoid building
momentum in staled directions.  For modeling $\mu$ and $\sigma^2$, we use
factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
$\gamma=5$.

\begin{figure}
    \centering
    \vspace{-0.8cm}
    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.}\label{fig:gaussian}
\end{figure}
\begin{figure}
    \centering
    \vspace{-0.5cm}
    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on discrete data: Poisson with unknown scale.}\label{fig:poisson}
\end{figure}

The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
after the optimization procedure. As theoretically expected from adversarial training, the
model density $p(\mathbf{x}|\psi)$ closely matches the true density
$p_r(\mathbf{x})$, as shown by the aligned histograms in blue and orange. Under
the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we also observe
that the proposal distributions have correctly concentrated their mass around
$\mu=1$ and $\log \sigma=-0.69$. The bottom plot of Figure~\ref{fig:gaussian}
shows an empirical estimate of $-U_d$ with respect to the epoch. We observe
that it quickly falls towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)} [d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$
and that the critic cannot distinguish between true and model data.

Similarly, we evaluate adversarial variational optimization on discrete data,
for the inference of the scale parameter $\lambda$ of a Poisson distribution.
The observed data is sampled from a univariate Poisson with scale $\lambda=7$.
We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
initialized with a mean at $5$ and unit variance. Hyper-parameters are otherwise
left unchanged with respect to the previous setup. Despite the discreteness
and the non-differentiability of the underlying generator, Figure~\ref{fig:poisson}
shows that inference with adversarial variational optimization works.
The upper plot in the figure shows that $p(\mathbf{x}|\psi)$ closely
matches $p_r(\mathbf{x})$, with the mass of the proposal distribution
concentrated at $\log 7 = 1.94$.

@@

>>>>

In this note, we develop a likelihood-free inference algorithm for the point
estimation of the parameters of a forward non-differentiable
generative model. We adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an arbitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

Variational optimization~\cite{2012arXiv1212.4507S} and evolution strategies~\citep{2011arXiv1106.4487W} are general
optimization techniques that can be used to form a differentiable bound
on the optima of a non-differentiable function. Given a function $f$ to minimize,
these techniques are based on the simple fact that
\begin{equation}
    \min_{\theta \in \Theta} f(\theta) \leq \mathbb{E}_{\theta \sim q(\theta|\psi)} [f(\theta)] = U(\psi),
\end{equation}
where $q(\theta|\psi)$ is a proposal distribution with parameters $\psi$ over input values $\theta$.
That is, the minimum of a set of function values is always less than or equal
to any of their average. Provided that the proposal is flexible enough, the parameters $\psi$
can be updated to place its mass arbitrarily tight around the optimum $\theta^* = \min_{\theta \in \Theta} f(\theta)$.

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q(\theta|\psi)$ over the
parameters of $g$ and $p(\mathbf{x}|\theta)$ and minimize in alternation the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$\phi$, unbiased estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$\psi$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
That is,
\begin{equation}
    \nabla_\psi U_g = \mathbb{E}_{\theta \sim q(\theta|\psi), \mathbf{z} \sim p(\mathbf{z}|\theta)}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q(\theta|\psi)],
\end{equation} which we can approximate with mini-batches of
generated data
\begin{equation}
    \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q(\theta_m|\psi)
\end{equation}
for $\theta_m \sim q(\theta|\psi)$ and $\mathbf{z}_m \sim p(\mathbf{z}|\theta_m)$ and .
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}
For completeness, Algorithm~\ref{alg:avo} outlines the proposed adversarial variational
optimization procedure, as built on top of WGAN-GP.
Obviously, the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

@@

In consequence, the proposed inference algorithm does not necessarily guarantee that the
proposal distribution $q(\theta|\psi)i$ will place its mass arbitrarily tight
around the parameters of interest, which might be an issue when one is rather interested in point estimates $\theta^*$.
For this purpose, we augment Eqn.~\ref{eqn:vo-ug}
with a regularization term corresponding to the differential entropy $H$ of
the proposal distribution. That is,
\begin{equation}
    U_g = \mathbb{E}_{\theta \sim q(\theta|\psi)} [ {\cal L}_g ] + \gamma H(q(\theta|\psi))
\end{equation}
where $\gamma \in \mathbb{R}^+$ is a hyper-parameter controlling the trade-off
between the generator objective and the tightness of the proposal distribution.
For small values of $\gamma$,
proposal distributions with large entropy are not penalized, which may result
in learning a smeared variation of the original simulator. On the other hand,
for large values of $\gamma$, the procedure is constrained to fit a proposal
distribution with low entropy, which has the effect of concentrating its density
tightly around one or a few $\theta$ values. A too large penalty may however
eventually make the optimization unstable, as the variance of $\nabla_\psi \log q(\theta_m|\psi)$
typically increases as the entropy of the proposal decreases.

@@

For continuous data, we first consider inference for
a univariate Gaussian with unknown mean and
unknown variance. The observed data is sampled from a univariate Gaussian with
mean $\mu^* = 1$ and standard deviation ${\sigma^*} = 0.5$. Algorithm~\ref{alg:avo} is run
for 300 epochs with mini-batches of size $M=64$ and the following configuration.
For the critic, we use a 3-layer MLP with 10 hidden nodes per layer and ReLU
activations. At each epoch, Adam is run for $n_\text{critic}=100$ iterations
with a step size $\alpha=0.01$, decay rates $\beta_1=\beta_2=0.5$ and its inner
first and second moment vectors reset at each epoch in order to avoid building
momentum in staled directions.  For estimating $\theta^* = (\mu^*, \sigma^*)$, we use
factored univariate Gaussian proposal distributions $q(\mu|\psi)$ and $q(\log
\sigma|\psi)$ initialized with zero mean and unit variance. At each epoch,
parameters $\psi$ are updated by making one Adam step, with $\alpha=0.01$ and
$\beta_1=\beta_2=0.1$. Penalty coefficients are set to $\lambda=0.025$ and
$\gamma=5$.

\begin{figure}
    \centering
    \vspace{-0.8cm}
    \includegraphics[width=0.48\textwidth]{figures/gaussian-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on continuous data: Gaussian with unknown mean and unknown variance.
             (Top) Model distribution $p(\mathbf{x}|\psi)$ after adversarial variational optimization along with the proposal distributions $q(\mu|\psi)$ and $q(\log
             \sigma|\psi)$ over the generator parameters.
             (Bottom) Evolution of the variational upper bound $U_d$ on the loss ${\cal L}_d$ of the critic. }\label{fig:gaussian}
\end{figure}
\begin{figure}
    \centering
    \vspace{-0.5cm}
    \includegraphics[width=0.48\textwidth]{figures/poisson-gamma=5.pdf}
    \vspace{-1cm}
    \caption{Toy problem on discrete data: Poisson with unknown scale.
             This example illustrates the effectiveness of adversarial variational optimization on discrete non-differentiable generators.}\label{fig:poisson}
\end{figure}

The upper plot of Figure~\ref{fig:gaussian} illustrates the resulting model
after the optimization procedure. As theoretically expected from adversarial
training, the model density $p(\mathbf{x}|\psi)$ closely matches the true
density $p_r(\mathbf{x})$, as shown by the aligned histograms in blue and
orange. Under the effect of the entropy penalty $\gamma H(q(\theta|\psi))$, we
also observe that the proposal distributions have correctly concentrated their
mass around $\mu^*=1$ and $\log \sigma^*=-0.69$. The bottom plot of
Figure~\ref{fig:gaussian} shows an empirical estimate of $-U_d$ with respect to
the epoch. We observe that it quickly falls towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\theta)}
[d(\tilde{\mathbf{x}};\phi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\phi)]$ and that the critic cannot distinguish
between true and model data.

Similarly, we evaluate adversarial variational optimization on discrete data,
for the inference of the scale parameter $\lambda$ of a Poisson distribution.
The observed data is sampled from a univariate Poisson with scale $\lambda^*=7$.
We use a univariate Gaussian proposal distribution $q(\log \lambda|\psi)$
initialized with a mean at $\log 5$ and unit variance. The optimization setup is
otherwise left unchanged. Despite the discreteness and the non-differentiability
of the underlying generator, Figure~\ref{fig:poisson} shows that inference with
adversarial variational optimization works. The upper plot in the figure shows
that $p(\mathbf{x}|\psi)$ closely matches $p_r(\mathbf{x})$, with the mass of
the proposal distribution concentrated at $\log 7 = 1.94$.

@@

\glnote{avo = extension of gans to non-differentiable generator}
\glnote{nice for discrete data}
